% ============================================================
%  Evaluation section for LectureForge.
%  \input{evaluation_section} into your main document.
%  Requires (in the preamble):
%     \usepackage{booktabs}
%     \usepackage{amsmath}
%  The prompt is shown with the built-in `verbatim` environment, so no extra
%  package is needed. (If you prefer a framed/coloured box, load `listings` and
%  swap the verbatim block for lstlisting.)
% ============================================================

\section{Evaluation}
\label{sec:evaluation}

We evaluate the question generator by comparing two pipelines on the same lecture
and scoring their output with an independent ``LLM-as-judge''. The goal is to test
whether the agentic pipeline produces better quiz questions than a non-agentic
baseline, and to quantify \emph{how} they differ.

\subsection{Protocol and models}

For a given quiz type (MCQ, true/false, fill-in-the-blank, short answer) we run:
\begin{itemize}
  \item \textbf{Agentic pipeline} --- a per-question loop of
        \emph{Plan $\rightarrow$ Generate $\rightarrow$ Validate $\rightarrow$ Retry}.
        Each question is drafted, checked by a self-validator on grounding,
        difficulty, instruction-compliance and uniqueness, and regenerated (with
        the failure reason fed back) until it passes or a retry budget is reached.
        The generator may also issue its own retrieval query
        (\texttt{search\_lecture}) over the lecture index when it judges the
        supplied context insufficient.
  \item \textbf{Non-agentic baseline} --- a single LLM call that writes one
        question on the same topic, with no validation, no retrieval and no retry.
\end{itemize}

Both pipelines generate from identical lecture content (transcript / slide OCR /
document text, packed the same way). The generator and self-validator use a fast
model (\texttt{gemma3:4b}); the judge uses a larger, separate model
(\texttt{gemma3:12b}) so that the model scoring the questions never wrote or
validated them, reducing self-grading bias. All inference is local (Ollama) with
JSON-constrained decoding.

\subsection{Per-question scoring: the judge}

Every generated question (from both pipelines) is sent individually to the judge
with the grounding content and a strict rubric. The judge returns a 1--5 score on
four dimensions plus a binary accept/reject decision:

\begin{verbatim}
You are an impartial, strict but fair exam-quality judge. Score this ONE
[TYPE] question on a 1-5 scale for each dimension.

Target difficulty: [DIFF] - [difficulty description]
Content snippet (ground truth):
[up to 4500 chars of lecture content]

Question:
Q: [question text]
Options: [options]          (MCQ only)
A: [answer]

Score 1 (worst) to 5 (best) on:
1. correctness        - factually correct, answerable from the content
2. clarity            - unambiguous, well-formed wording
3. difficulty_match   - actual difficulty matches the target
4. distractor_quality - wrong options plausible but clearly wrong (MCQ only)
5. overall            - holistic quality

BE DISCRIMINATING - use the FULL range, do not default to 5:
  5 = flawless; 4 = good; 3 = ordinary; 2 = weak; 1 = poor.
Reserve 5 for genuinely excellent questions; penalise generic phrasing,
near-duplicate framing, weak distractors, shallow recall.
Then decide accept (true) or reject (false).

Return STRICT JSON:
{"correctness":1,"clarity":1,"difficulty_match":1,
 "distractor_quality":1,"overall":1,"accept":true,"reason":""}
\end{verbatim}

The judge fails closed (a parse error yields a rejected, zero-scored verdict, so
failures cannot inflate the result), and \texttt{overall}$<3$ forces
\texttt{accept}$=$\texttt{false} for consistency. The agentic self-validator uses a
parallel but lighter prompt that returns
\texttt{\{difficulty, grounded, instruction, overall\}} as
\texttt{PASS}/\texttt{FAIL} together with a remediation action
(\texttt{regenerate}, \texttt{fetch\_context}, or \texttt{replan\_topic}).

\subsection{Metrics}

\paragraph{Evaluation procedure.} For each quiz type and target difficulty
$d\in\{\text{easy},\text{medium},\text{hard}\}$ we evaluate as follows. Let
$Q^{A}=\{q^{A}_1,\dots,q^{A}_N\}$ be the questions produced by the agentic pipeline
and, for each agentic question's topic $t_i$, let $q^{B}_i$ be the question the
non-agentic baseline produces on that same topic, giving $Q^{B}=\{q^{B}_1,\dots,q^{B}_N\}$.
Both sets are conditioned on identical lecture content $C$. Every question is then
scored \emph{independently} by the judge model $J$, which returns
$J(q)=(\text{correctness},\text{clarity},\text{difficulty\_match},\text{overall},\text{accept})$
with each numeric field in $[1,5]$ and $\text{accept}\in\{0,1\}$. We aggregate the
per-question judge outputs into set-level means, combine them with the
deterministic and embedding metrics below, and then rank the two pipelines
(Sec.~``Ranking''). Each configuration is repeated $R$ times and all reported
figures are means over the $N\!\times\!R$ questions.

Each question set is thus measured along three metric families: judge-based,
deterministic, and embedding-based.

\paragraph{Judge metrics.} Mean overall/correctness/clarity/difficulty-match
(1--5) and the \emph{accept rate} $\mathrm{Acc}=\frac{1}{N}\sum_i \text{accept}(q_i)$
(fraction the judge would put on a real exam).

\paragraph{Deterministic metrics.} A \emph{duplicate rate} flags near-identical
questions using a string-similarity ratio with threshold $0.85$; we also report
structural validity (e.g.\ the MCQ answer key is a real option, true/false balance,
presence of the blank marker).

\paragraph{Embedding metrics.} Using the sentence-embedding model that powers
retrieval, for a question $q_i$ and lecture chunks $\{c_j\}$ we define
\emph{grounding} as the mean best-match similarity and \emph{diversity} as one
minus the mean nearest-neighbour similarity between questions:
\begin{equation}
  \mathrm{Grounding} = \frac{1}{N}\sum_{i=1}^{N}\max_{j}\cos\!\big(\mathbf{e}(q_i),\mathbf{e}(c_j)\big),
\end{equation}
\begin{equation}
  \mathrm{Diversity} = 1 - \frac{1}{N}\sum_{i=1}^{N}\max_{k\neq i}\cos\!\big(\mathbf{e}(q_i),\mathbf{e}(q_k)\big).
\end{equation}

\paragraph{Validator--judge agreement (agentic only).} We treat the independent
judge's \texttt{accept} as ground truth and the agentic self-validator's
\texttt{PASS} as the prediction, and report precision, recall and~$F_1$:
\begin{equation}
  P = \frac{TP}{TP+FP}, \quad
  R = \frac{TP}{TP+FN}, \quad
  F_1 = \frac{2PR}{P+R}.
\end{equation}
This measures whether the agent's own quality gate matches an external judge.

\subsection{Ranking}

We rank the two pipelines at two levels: per dimension (a \emph{segment verdict})
and overall.

\paragraph{Per-segment ranking.} The judge's per-question scores are averaged into
three segments --- \emph{overall quality}, \emph{in-context/grounded}, and
\emph{difficulty match}. For a pipeline $X$ and segment $s$, the segment score is
\begin{equation}
  S_s(X) = \frac{1}{N}\sum_{i=1}^{N} \mathrm{judge}_s\!\big(q^{X}_i\big),
\end{equation}
and the winner of segment $s$ is the pipeline with the larger $S_s$. Because the
judge scores saturate (good questions cluster at the top of the $1$--$5$ scale),
exact ties are common; we therefore break a tie deterministically with a
content-quality margin that rewards grounding and penalises repetition,
\begin{equation}
  T(X) = \mathrm{Grounding}(X) - \mathrm{DuplicateRate}(X),
\end{equation}
awarding the tied segment to the pipeline with the larger $T$.

\paragraph{Overall ranking (duplicate-penalised).} A question that merely repeats
an earlier one carries little exam value, yet the judge --- scoring each question
in isolation --- cannot see this. To rank at the set level we therefore use an
\emph{effective accept rate} that discounts the judge's accept rate by the
fraction of redundant questions,
\begin{equation}
  \mathrm{Acc}_{\mathrm{eff}}(X) = \mathrm{Acc}(X)\,\big(1-\mathrm{DuplicateRate}(X)\big),
\end{equation}
so a set of distinct, accepted questions scores highest while a repetitive set is
down-weighted in proportion to its redundancy. The pipeline with the larger
$\mathrm{Acc}_{\mathrm{eff}}$ is ranked higher overall; we report
$\mathrm{Acc}$, $\mathrm{DuplicateRate}$ and $\mathrm{Acc}_{\mathrm{eff}}$ together
so the effect of repetition on the ranking is transparent.

\subsection{Findings}

\begin{table}[t]
\centering
\small
\begin{tabular}{lcccc}
\toprule
 & \multicolumn{2}{c}{Medium, $N{=}10$} & \multicolumn{2}{c}{Hard, $N{=}5$} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
Metric & Agentic & Non-agentic & Agentic & Non-agentic \\
\midrule
Overall (judge, 1--5)   & 5.0  & 5.0  & 3.6  & 5.0 \\
In-context (1--5)        & 5.0  & 5.0  & 3.4  & 5.0 \\
Difficulty match (1--5)  & 5.0  & 5.0  & 3.2  & 4.0 \\
Accept rate              & 100\% & 100\% & 60\% & 100\% \\
Grounding                & 0.569 & 0.637 & 0.453 & 0.612 \\
Duplicate rate           & 30\% & 40\% & \textbf{0\%} & \textbf{80\%} \\
Validator--judge $F_1$   & 0.571 & --- & 0.0\textsuperscript{*} & --- \\
\bottomrule
\end{tabular}
\caption{Agentic vs.\ non-agentic on one lecture. \textsuperscript{*}$F_1$ is
unreliable at small $N$.}
\label{tab:eval}
\end{table}

Three observations emerge (Table~\ref{tab:eval}):

\begin{enumerate}
  \item \textbf{The judge score alone is insufficient.} At medium difficulty the
        judge assigned $5/5$ to nearly every question on both sides, so all
        segments tied; the score saturates and cannot separate the pipelines. This
        motivated (a) recalibrating the judge to use the full range and (b) a
        deterministic tiebreaker.
  \item \textbf{Diversity is the true discriminator.} The duplicate rate
        consistently separated the pipelines: the agentic pipeline was far less
        repetitive (0--30\%) than the baseline (40--80\%). At hard difficulty the
        baseline scored $5/5$ only by asking essentially the \emph{same} question
        repeatedly ($80\%$ duplicates), a weakness invisible to a per-question
        judge but captured by the metric.
  \item \textbf{Agentic generation trades safety for ambition.} At hard
        difficulty the agentic pipeline attempted genuinely harder questions but
        occasionally introduced content absent from the lecture (e.g.\ net present
        value), which the judge correctly rejected, lowering its grounding and
        accept rate. Constraining the generator to remain answerable from the
        lecture material addresses this. The validator--judge $F_1$ ($P{=}1.0$,
        $R{=}0.4$ at $N{=}10$) further shows the agent's self-validator is
        conservative --- it never approves a question the judge would reject, but
        rejects many the judge would accept.
\end{enumerate}

\paragraph{Limitations.} The judge is an LLM rather than a human rater, so absolute
scores are a relative signal; generator and judge share the Gemma-3 family
(mitigated, not eliminated, by using a larger independent judge); and results are
from a single lecture, so $F_1$ and per-segment verdicts require $N\gtrsim10$ and
multiple lectures to be statistically meaningful.
